---
title: "Flow_Analysis_tutorial_2024"
author: "Ben Barry"
date: "2024-11-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install packages
```{r installing the required packages}
#You can either do library(BiocManager) and install("flowCore) or what I have done below.
BiocManager::install("flowCore") #interpret .fcs files
BiocManager::install("flowViz") #basic visulisation
BiocManager::install("ggcyto") #advanced visulisation using the ggPlot nomenclature
BiocManager::install("openCyto") #Used to link various analysis methodologies
BiocManager::install("flowWorkspace") #used to build anaysis templates
BiocManager::install("CytoML") #imports FlowJo and DiVA workspaces

#These packages largly require each other to work (except flowCore which is the "base package) 
#so will often load each other without my help.  For simplicty I have loaded them all.

#You will need to "clean" your data.  flowAI and flowCut are my recomendations.  
#flowClean is the original, but is supeceeded by flowCut
BiocManager::install("flowAI")
BiocManager::install("flowClean")
BiocManager::install("flowCut")

#The CytoExploreR package is not available on bioconductor and needs to be loaded straight from GitHub.  To do this you need the package devtools.
install.packages("devtools")

#An intresting project is CytoExploreR that trys to blend the power of R with the ease of use of a mouse.
devtools::install_github("DillonHammill/CytoExploreRData")
devtools::install_github("DillonHammill/CytoExploreR", build_vignettes = FALSE)

```

# Part 1: Downloading the Flow Dataset.
I was unable to download data from FlowRepository, however I found some data from a different tutorial
https://jchellmuth.com/posts/FACS-with-R/
this data is very simple only measureing desinged only to measure change in GFP (FITC-A) - however it serves the same for this pupose. 

```{r download Flow dataset}
FlowFiles <- list.files(
  path = "/Users/benbarry/Documents/Flow_Analysis_tutorial_2024/data/2020-07-08-FACS-data/",
  pattern = ".fcs$"
)

FlowSet <- flowCore::read.flowSet(FlowFiles, path = "/Users/benbarry/Documents/Flow_Analysis_tutorial_2024/data/2020-07-08-FACS-data/")

FlowSet
FlowSet[[1]] #view the first FCS files
```

# Part II: compensation cleaning and transforming
## Compensation:
Corrects for spectral overlap the method here uses a compensation matrix - however there since this is a simple dataset
the compensation matrix is very simple. 

## Cleaning:
here cleaning refers to a quick QC check 
It consists in the detection and removal of anomalies by checking three properties of flow cytometry: 1) flow rate, 2) signal acquisition, 3) dynamic range.

## Transfomring:
here we are transforming the data to make it more interpertable using a logical transform. Logicle transformation creates a subset of biexponentialTransform hyperbolic sine transformation functions that provides several advantages over linear/log transformations for display of flow cytometry data.

```{r compensation, cleaning, tranfomration of flow data}
library(flowCore)
library(flowAI)
library(ggcyto)

#load a single file 
singleFile <- "/Users/benbarry/Documents/Flow_Analysis_tutorial_2024/data/2020-07-08-FACS-data/Specimen_001_B2_B02_001.fcs"

fcsFile <- read.FCS(singleFile)
fcsFile
names(fcsFile)
exprs(fcsFile)
each_col(fcsFile, median)
keyword(fcsFile)

#compensation
spillover(fcsFile)
fcsFileComp <- compensate(fcsFile, spillover(fcsFile)$SPILL)

#cleaning the data
fcsFileCompClean <- flow_auto_qc(fcsFileComp)
fcsFileCompClean

#transforming the data 
trans <- estimateLogicle(fcsFileCompClean, colnames(fcsFileCompClean[, 7:8]))

fcsFileCompCleanTrans <- transform(fcsFileCompClean, trans)


#visualise the results
autoplot(fcsFileCompCleanTrans)
autoplot(fcsFileCompClean)
autoplot(fcsFileCompCleanTrans, x = "FITC-A", y = "Pacific Blue-A", bin = 200)
autoplot(fcsFileCompCleanTrans, x = "Time", y = "Pacific Blue-A", bin = 200)

#working with a full flowset.
FlowFiles <- list.files(
  path = "/Users/benbarry/Documents/Flow_Analysis_tutorial_2024/data/2020-07-08-FACS-data/",
  pattern = ".fcs$")

corrupted_files <- c("Specimen_001_C11_C11_020.fcs",
                     "Specimen_001_C9_C09_018.fcs", 
                     "Specimen_001_D11_D11_030.fcs")

FlowFiles <- FlowFiles[!FlowFiles %in% corrupted_files]

FlowSet <- flowCore::read.flowSet(FlowFiles, 
                                  path = "/Users/benbarry/Documents/Flow_Analysis_tutorial_2024/data/2020-07-08-FACS-data/")

FlowSet 
spillover(FlowSet[[1]])
FlowSet_Comp <- compensate(FlowSet, spillover(FlowSet[[1]])$SPILL)
FlowSet_Comp_Clean <- flow_auto_qc(FlowSet_Comp) #several files seemed to be corrupted these were removed. 
FlowSet_Trans <- estimateLogicle(FlowSet_Comp_Clean[[1]], colnames(FlowSet_Comp_Clean[[1]][, 7:8]))
FlowSet_Comp_Clean_Trans <- transform(FlowSet_Comp_Clean, FlowSet_Trans)
autoplot(FlowSet_Comp_Clean_Trans[[1]] , x = "FITC-A", y = "Pacific Blue-A", bin = 200)

```

# Manual gating
## Part III: Gating - Manual 

Cell gating isolates cells based on forward and side scatter (FSC-A vs. SSC-A) to remove debris. Singlet gating is then applied using FSC-H vs. FSC-W to exclude doublets or cell clumps. The `gs_get_pop_paths()` function shows how data flows from the cell gate to the singlet gate, applying each filter sequentially.

```{r gating - manual and auto}
#plotting here with sample 13
autoplot(FlowSet_Comp_Clean_Trans[[13]])

ggcyto(FlowSet_Comp_Clean_Trans[[13]], 
       aes(x = "FSC-A",
           y = "SSC-A")) + 
  geom_hex(bins = 200)

gs <- GatingSet(FlowSet_Comp_Clean_Trans) #create an empty gating set 

# gating cells 
rectangle <- rectangleGate("FSC-A" = c(75000, Inf),
                           filterId = "NoneDebris")

gs_pop_add(gs, rectangle, parent = "root")
recompute(gs)
gs_get_pop_paths(gs)

ggcyto(FlowSet_Comp_Clean_Trans[[13]], 
       aes(x = "FSC-A",
           y = "SSC-A")) + 
  geom_hex(bins = 200) + 
  geom_gate(gs_pop_get_gate(gs, "NoneDebris"))

gs_pop_get_stats(gs)

# singlet gating

ggcyto(FlowSet_Comp_Clean_Trans[[13]], 
       aes(x = "FSC-H",
           y = "FSC-W")) + 
  geom_hex(bins = 200) +
  scale_x_continuous(n.breaks = 8)

rec2 <- rectangleGate("FSC-H"=c(2e4, 1.1e5),"FSC-W"=c(5e4, 1.25e5))
gs_pop_add(gs, rec2, parent = "NoneDebris", name = "singlets")
gs_get_pop_paths(gs)
recompute(gs)

ggcyto(FlowSet_Comp_Clean_Trans[[25]], 
       aes(x = "FSC-H",
           y = "FSC-W")) + 
  geom_hex(bins = 200) + 
  geom_gate(gs_pop_get_gate(gs, "singlets"))

## exploring the gating set

plot(gs)
gs_pop_get_stats(gs)
gs_pop_get_stats(gs, "NoneDebris", "percent")
```

## Part IV: Gating - Auto

This section automates the gating process using the `openCyto` package. The `noneDebrisGate` is applied using a flowClust 2D method to remove debris, based on FSC-A vs. SSC-A, which is similar to the manual gating approach but now automated across all samples. Next, singlet gating is performed automatically using FSC-A vs. FSC-H to exclude doublets, similar to the manual approach but applied in an automated manner using `fsApply`. Next we apply quad gating on two channels (FITC-A vs. Pacific Blue-A), which further subdivides the data into four populations. one limitation here is that really the `noneDebrisGate` should be set on a control sample and applied across all samples. 

###  Part IV: Gating - Auto basic Stats
The `pop.quantiles()` function calculates the 75th percentile (Q3) for each channel in the dataset, providing a measure of the upper distribution for each population. The `pop.mean()` function computes the mean expression value for each channel across the gated population, offering an average intensity measure for each channel. These functions are applied to each gated population using `gs_pop_get_stats()`, which helps summarize the data by calculating these statistical metrics across the different gates.


```{r auto gating}
library(openCyto)
AutoGate <- GatingSet(FlowSet_Comp_Clean_Trans)


#cell gate - removing debris

Flowset_Data <- gs_pop_get_data(AutoGate)
noneDebrisGate <- fsApply(Flowset_Data, function(fr) openCyto::gate_flowclust_2d(fr, xChannel = "FSC-A", yChannel = "SSC-A"))
gs_pop_add(AutoGate, noneDebrisGate, parent = "root", name = "noneDebrisGate")
recompute(AutoGate)
autoplot(AutoGate[[13]], x = "FSC-A", y = "SSC-A", "noneDebrisGate",  bins = 200)

#Singlet gating 

Flowset_Data <- gs_pop_get_data(AutoGate, "noneDebrisGate")
singletGate <- fsApply(Flowset_Data, function(fr) openCyto:::.singletGate(fr, channels = c("FSC-A","FSC-H")))

gs_pop_add(AutoGate, singletGate, parent = "noneDebrisGate", name = "singlets")
recompute(AutoGate)
autoplot(AutoGate[[1]], x = 'FSC.A', y = 'FSC.H', "singlets", bins = 256)

plot(AutoGate)
gs_pop_get_stats(AutoGate)
gs_pop_get_stats(AutoGate, "noneDebrisGate", "percent")


#Quad gating. 
Flowset_Data <- gs_pop_get_data(AutoGate, "singlets") 
gateQuad <- fsApply(Flowset_Data, function(fr) gate_quad_sequential(fr, channels = c("FITC-A", "Pacific Blue-A"), gFunc = "mindensity"))
gs_pop_add(AutoGate, gateQuad, parent = "singlets", names = c("1", "2", "3", "4"))
recompute(AutoGate)
gs_get_pop_paths(AutoGate[[1]])
plot(AutoGate)
autoplot(AutoGate[[13]], x = "FITC-A", y = "Pacific Blue-A", gs_get_pop_paths(AutoGate)[4:7], bins = 150)


gs_pop_get_stats(AutoGate)
gs_pop_get_stats(AutoGate, "noneDebrisGate", "percent")

pop.quantiles <- function(fr){
  channels = colnames(fr)
  res <- matrixStats::colQuantiles(exprs(fr), probs = 0.75)
  names(res) <- channels
  res
}

gs_pop_get_stats(AutoGate, gs_get_pop_paths(AutoGate), type = pop.quantiles)

pop.mean <- function(fr){
  channels <- colnames(fr)
  res <- colMeans(exprs(fr))
  names(res) <- channels
  res
}
gs_pop_get_stats(AutoGate, gs_get_pop_paths(AutoGate), type = pop.mean)

```